### 2018-12-13 submission/manuscript.Rmd
Edit results section

**Background.** A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, amplification, and sequencing steps. At each of these stages error can be introduced that will change the underlying bacterial community composition results. In this study, we focus on the amplification step's contribution to this overall error. To accomplish this we assessed 16S rRNA gene sequencing results in mock community and human fecal samples after using different  amplification cycles and high fidelity DNA polymerases.

**Methods.** Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria were amplifeid at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 high fidelity DNA polymerase. For fecal samples (n=4) we first extracted DNA using a PowerMag DNA extraction kit with a 10 minute bead beating step and then used the same approach as the mock community for the PCR amplification step. We first assessed GC dependent differences, error rate, sequence error prevalence, chimera prevalence, and correlation between chimera prevalence and number of Operational Taxonomic Units (OTUs) by polymerase and number of cycles. Next, differences in the number of OTUs was examined based on the polymerase and number of cycles used. Additionally, differences in the bacterial community composition was assessed based on polymerase and number of cycles using the Bray-Curtis index. Based on these findings individual taxa differences based on polymerase and number of cycles was also investigated. Finally, Random Forest models were created to test whether the bacterial community was better at classifying polymerases, number of cycles, or individual donor. We also assessed whether the most important taxa in the polymerase and number of cycle Random Forest models were also important in the model for individual donors.

**Results.** Predictably, we found noticeable differences in relative abundance based on high and low GC content (P-value $\leqslant$ `r round(max(filter(gc_data, bh < 0.05) %>% pull(bh)), digits = 2)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2^ = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2^ = `r min(pull(r2_table, r2))`). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). When analyzing fecal samples we observed smaller differences in the number of OTUs between polymerases. The median number of OTUs only varied by high fidelity DNA polymerase used at 35 cycles and not at 20 cycles like the mock community (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Random Forest models were most successful at classifying individual donor samples rather than polymerase or number of cycles used (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal"), str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`). Additionally, the most important OTUs in the polymerase and number of cycle models were not the most important in the individual donor model.

**Conclusions.** Although there were 16S rRNA gene sequencing differences based on polymerase and number of cycles used, bacterial community composition differences were mostly only detectable in mock communities. Collectively, these results provide evidence that smaller biological differences between groups, based on 16S rRNA gene sequencing of human fecal samples, can be consistently detected regardless of polymerase and number of cycles used.

@@

***Sequencing errors vary by the number of cycles and the DNA polymerase used in PCR.*** The presence of sequence errors can confound the ability to accurately classify 16S rRNA gene sequences and group sequences into operational taxonomic units (OTUs). More importantly, sequencing errors themselves can alter the representation of the community. Therefore, it is important to minimize the number of sequencing errors in an analysis. Using a widely-used approach that generates the lowest reported error rate, we quantified the error rate by sequencing the V4 region of the 16S rRNA genes from an 8 member mock community. We also removed any contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence [ref]. Regardless of the polymerase, the error rate increased with the number of rounds of amplification. Using 30 rounds of PCR is a common approach across diverse types of samples. Among the data generated using 30 rounds of PCR the Accuprime polymerase had the highest error rate (i.e. 0.XXX%) followed by the Platinum (i.e. 0.XXX%), Phusion (i.e. 0.XXX%), Kappa (i.e. 0.XXX%), and Q5 (i.e. 0.XXX%) polymerases (Figure 1). The order of the error rates using these polymerases is consistent with previously documented reports of their per cycle error rates. When we applied the pre.clustering denoising step, which merges the counts of reads within 2 nt of a more abundant sequence, the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e. 0.XXX%) followed by the Accuprime (i.e. 0.XXX%), Phusion (i.e. 0.XXX%), Kappa (i.e. 0.XXX%), and Q5 (i.e. 0.XXX%) polymerases (Figure 1). Although specific recommendations are difficult to make because the impact of the actual community structure and concentration of the initial DNA template are likely to have an impact on the results, it is clear that using as few PCR cycles as necessary is a good guide to minimizing the impact of polymerase on the error rate. Furthermore, using a polymerase with the lowest possible error rate is preferred.

***The fraction of sequences identified as being chimeric varies by the number of cycles and the DNA polymerase used in PCR.*** Chimeric PCR products can significantly confound downstream analyses. Although numerous bioinformatic tools exist to identify and remove chimeric sequences with high specificity, their sensitivity is relatively low and can be reduced by the presence of sequencing errors. We identified those contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence. As expected from previous studies, the number of chimeras increased with rounds of amplification. Interestingly, the fraction of chimeric sequences from the mock community varied by the type of polymerase used. After 30 rounds of PCR, the Platinum polymerase had the highest chimera rate (i.e. XX%) followed by the Q5 (i.e. XX%), Phusion (i.e. XX%), Kappa (i.e. XX%), and Accuprime (i.e. XX%) polymerases. We used the UChime algorithm to detect chimeras in our pre-clustered mock community data and calculate the algorithm's sensitivity and specificity. For all polymerases, the specificity was above XX% and showed a slight association with the number of cycles used. There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UChime to detect the chimeras from the mock community. This suggests that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity. In general, the sensitivity of UChime varied between 50 and 90%. These results could be affected by the use of a single mock community with limited genetic diversity. Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UChime identified as being chimeric. These results followed those from the mock communities - additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used. Although it is difficult to identify a feature of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.

***The bias of PCR amplification is inconsistent and subtle.*** Since researchers began using PCR to sequence 16S rRNA genes there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses. The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria. The vendor of the mock community subjects each lot of genomic DNA to shotgun sequencing to more accurately quantify the actual abundance of each organism in the community. We compared the expected relative abundance of the 16S rRNA genes from each bacterium in the mock community to the data we generated across rounds of amplification and polymerase. Interestingly, for some bacteria, there representation became less biased with additional rounds of PCR (e.g. *L. fermentum*), while others became more biased (e.g. *E. faecalis*), and others had little change (e.g. *B. subtilis*). The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias. Across the bacteria there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed. One of the bacteria represented in the mock community, *S. enterica*, had 6 identical copies of the V4 region and 1 copy that differed from those by one nucleotide. We used the sequence data to calculate the ratio of the dominant to rare variants from *S. enterica* (Figure. XXX). With the exception of the Kappa polymerase, the other four polymerases converged to a ratio of 5.XXX. Based on the relative abundance data from the mock community it was not possible to discern the factors that result in amplification bias. As an additional test, we quantified the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification. For the mock community data, the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of 0.XX units. The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX). In contrast, when we characterized the variation between amplification rounds using human stool samples, the distance between the 25 and 30 rounds and 25 and 35 rounds varied considerably between samples and polymerases. In general the inter-round variation was lowest for the data generated using the Kappa and Accuprime polymerases. The data generated using the Platinum polymerase was consistent across rounds, but it was more biased than these polymerases. Considering the average distance across the four samples varied between 0.XX and 0.XX, regardless of the polymerases and number of rounds of amplification, any bias due to amplification is unlikely to obscure community-wide differences between samples. It is difficult to develop a specific recommendation based on the level of bias across rounds of PCR or polymerases; however, the general suggestion is to use as few rounds of amplification as possible.

@@

***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels appeared to be small relative to the overall differences between samples. To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification. We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor. Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions. We also performed a principle coordinates analysis of the Bray-Curtis distances showed distinct clusters by donor. Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR. However, when we performed the same analysis using the mock community, there were clear shifts in the represented community structures as the number of rounds of PCR increased. Using adonis, there were statistically significant differences in the community structures for the mock and stool community data (all P<0.001). Within the mock community data, the number of rounds of PCR explained the most variation (R^2^=0.XXX) followed by the choice of polymerase (R^2^=0.XXX). Similarly, within the stool community data, the subject explained the most variation (R^2^=0.XXX), followed by the number of rounds of PCR (R^2^=0.XXX) and the choice of polymerase (R^2^=0.XXX). Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples.

>>>>

@@

***Sequencing errors vary by the number of cycles and the DNA polymerase used in PCR.*** The presence of sequence errors can confound the ability to accurately classify 16S rRNA gene sequences and group sequences into operational taxonomic units (OTUs). More importantly, sequencing errors themselves can alter the representation of the community. Therefore, it is important to minimize the number of sequencing errors. Using a widely-used approach that generates the lowest reported error rate, we quantified the error rate by sequencing the V4 region of the 16S rRNA genes from an 8 member mock community. We also removed any contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence [ref]. Regardless of the polymerase, the error rate increased with the number of rounds of amplification. Using 30 rounds of PCR is a common approach across diverse types of samples. Among the data generated using 30 rounds of PCR the Accuprime polymerase had the highest error rate (i.e. 0.XXX%) followed by the Platinum (i.e. 0.XXX%), Phusion (i.e. 0.XXX%), Kappa (i.e. 0.XXX%), and Q5 (i.e. 0.XXX%) polymerases (Figure 1). When we applied the pre.clustering denoising step, which merges the counts of reads within 2 nt of a more abundant sequence, the error rates dropped considerably such that the Platinum polymerase had the highest error rate (i.e. 0.XXX%) followed by the Accuprime (i.e. 0.XXX%), Phusion (i.e. 0.XXX%), Kappa (i.e. 0.XXX%), and Q5 (i.e. 0.XXX%) polymerases (Figure 1). Although specific recommendations are difficult to make because the impact of the actual community structure and concentration of the initial DNA template are likely to have an impact on the results, it is clear that using as few PCR cycles as necessary and a polymerase with the lowest possible error rate is a good guide to minimizing the impact of polymerase on the error rate.

***The fraction of sequences identified as being chimeric varies by the number of cycles and the DNA polymerase used in PCR.*** Chimeric PCR products can significantly confound downstream analyses. Although numerous bioinformatic tools exist to identify and remove chimeric sequences with high specificity, their sensitivity is relatively low and can be reduced by the presence of sequencing errors. We identified those contigs that were at least three bases more similar to a chimera of two references than to a single reference sequence. As expected from previous studies, the number of chimeras increased with rounds of amplification. Interestingly, the fraction of chimeric sequences from the mock community varied by the type of polymerase used. After 30 rounds of PCR, the Platinum polymerase had the highest chimera rate (i.e. XX%) followed by the Q5 (i.e. XX%), Phusion (i.e. XX%), Kappa (i.e. XX%), and Accuprime (i.e. XX%) polymerases. We used the UChime algorithm to detect chimeras in our pre-clustered mock community data and calculate the algorithm's sensitivity and specificity. For all polymerases, the specificity was above XX% and showed a weak association with the number of cycles used. There was considerable inter-polymerase and inter-round of amplification variation in the sensitivity of UChime to detect the chimeras from the mock community. This suggested that the residual error rate after pre-clustering the sequence data did not compromise the sensitivity. In general, the sensitivity of UChime varied between XX and XX%. The generalizability of these results is limited because we used a single mock community with limited genetic diversity. Although we did not know the true chimera rate for our four human stool samples, we were able to calculate the fraction of sequences that UChime identified as being chimeric. These results followed those from the mock communities; additional rounds of amplification significantly increased the rate of chimeras and there was variation between the polymerases that we used. Although it was not possible to identify the features of a polymerase that resulted in higher rates of chimeras, it is clear that using the smallest number of PCR cycles possible will minimize the impacts of chimeras.

***At the sequence level, PCR amplification bias is subtle.*** Since researchers began using PCR to sequence 16S rRNA genes there has been concern that amplifying fragments from a mixed template pool could lead to a biased representation in the pool of products and would confound downstream analyses. The mock community was generated by mixing equal amounts of genomic DNA from each of the 8 bacteria resulting in uneven representation of the *rrn* operons across the bacteria. The vendor of the mock community subjects each lot of genomic DNA to shotgun sequencing to more accurately quantify the actual abundance of each organism in the community. We compared the expected relative abundance of the 16S rRNA genes from each bacterium in the mock community to the data we generated across rounds of amplification and polymerase. Interestingly, for some bacteria, their representation became less biased with additional rounds of PCR (e.g. *L. fermentum*), while others became more biased (e.g. *E. faecalis*), and others had little change (e.g. *B. subtilis*). The percentage of bases in the V4 region that were guanines or cytosines was not predictive of the amount of bias. Across the strains there was no variation in the length of their V4 regions and they each had the same sequence in the region that the primers annealed. One of the bacteria represented in the mock community, *S. enterica*, had 6 identical copies of the V4 region and 1 copy that differed from those by one nucleotide. The dominant copy had a thymidine and the rare copy had a guanine. We used the sequence data to calculate the ratio of the dominant to rare variants from *S. enterica* (Figure. XXX). The Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of 5.XXX; however, the ratio for the Kappa polymerase varied between 6.XXX and 7.XXX for 25 to 35 rounds of PCR. Given the subtle nature of the variation in the relative abundances of each 16S rRNA gene fragment, it was not possible to create generalizable rules that would explain the bias and the previous observation regarding the guanine and cytosine relative abundance was not observed with our data.

***At the community level, the effects of PCR amplification bias grow with additional rounds of PCR.*** Because the distance between samples and across rounds of PCR could be artificially inflated due to sequencing errors and chimeras, we used analyzed the alpha and beta diversity of the mock community data at different phases of the sequence curation pipeline. First, we removed the chimeras from the mock community data as described above and mapped the individual reads to the OTUs that the 16S rRNA gene fragments would cluster into if there were no sequencing errors. This gave us a community distribution that reflected the distribution following PCR without any artifacts. Alpha diversity metrics such as richness and Shannon and Inverse Simpson diversity metrics increased with the number of rounds of PCR for all polymerases except for the Kappa polymerase, where the richness remained constant but the diversity decreased. These data suggest that PCR had the effect of making the community distribution more even than it was originally. Next, we used the true sequence diversity, but removed chimeras based on mapping reads to all possible chimeras between reference sequences, and clustered the reads to OTUs. The same trends were observed as with perfect sequence data except that the richness and diversity metrics trended higher. Finally, we used the observed sequence data and used the UCHIME algorithm to identify chimeras. Again, we observed the same trends as with the other analyses except the richness and diversity metrics trended higher. These comparisons demonstrated that although the bias at the sequence level was subtle, PCR does have a bias that is exacerbated by errors and chimeras. When we measured the Bray-Curtis distance between the communities observed after 25 rounds of amplification and those at 30 and 35 rounds of amplification the distances between 25 and 35 rounds were higher than between 25 and 30 rounds for each of the polymerases by an average of 0.XX units. The Platinum polymerase varied the most across rounds of amplification (25 vs 30 rounds: 0.XX; 25 vs 35 rounds: 0.XX). Although the distances between samples were small, the ordination of these distances showed a clear change in community structure with increasing rounds of PCR. This observation was supported by our statistical analysis, which revealed that the number of rounds of PCR (R^2^=0.XX, P=0.XXXX) was a larger factor than the choice of polymerase (R^2^=0.XX, P=0.XXXX). These results demonstrate that subtle differences in relative abundances can have an impact on overall community structure. This variation underscores the importance of only comparing sequence data that have been generated using the same PCR conditions.

***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels using mock community data appeared to be small relative to the expected differences between biological samples. To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification. We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor. Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions. When we characterized the variation between rounds of amplification using human stool samples, the distance between the 25 and 30 rounds and 25 and 35 rounds varied considerably between samples and polymerases. In general the inter-round variation was lowest for the data generated using the Kappa and Accuprime polymerases. The data generated using the Platinum polymerase was consistent across rounds, but it was more biased than these polymerases. Considering the average distance across the four samples varied between 0.XX and 0.XX, regardless of the polymerases and number of rounds of amplification, any bias due to amplification is unlikely to obscure community-wide differences between samples. In support of this was our principle coordinates analysis of the Bray-Curtis distances, which revealed distinct clusters by donor. Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR. However, our statistical analysis using adonis revealed statistically significant differences in the community structures with the subject explaining the most variation (R^2^=0.XXX, P=0.XXXX), followed by the number of rounds of PCR (R^2^=0.XXX, P=0.XXXX) and the choice of polymerase (R^2^=0.XXX, P=0.XXXX). Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples. It is difficult to develop a specific recommendation based on the level of bias across rounds of PCR or polymerases; however, the general suggestion is to use as few rounds of amplification as possible.

@@
